---
title: "Concentration of CO2 in the Atmosphere Since 1958"
author: "Kadir Korkmaz"
date: "11/24/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In this document, I analysed the consantration of CO2 change in the Atmosphere since 1958. You can access the original version of the data set from [this](http://scrippsco2.ucsd.edu/data/atmospheric_co2/primary_mlo_co2_record) link. Dataset is updated continuosly and I worked on the snapshot of the dataset which is taken at 11/24/2018. Snapshot of the dataset will be available with this notebook.


## Loading Dataset

In Data folder we have a csv file contains weekly measurements of CO2 consantration.
CSV file contains some comments related to data so we skipped that part when we are loading the data. Also CSV file does not contain header in so after loading the dataset I added column names.

```{r load data}
#Loading data from CSV file
weekly_co2_consantration = read.csv("./data/weekly_in_situ_co2_mlo.csv", header = FALSE, skip=44)

#Adding header
colnames(weekly_co2_consantration)=c("date","consantration")

weekly_co2_consantration$date <- as.Date(weekly_co2_consantration$date,format="%Y-%m-%d")

```

## A First Look on Datasets

Weekly measurement dataset contains only two columns. Explanation of the columns values are given on the comment section of the csv file as following : 

>The data file below contains 2 columns indicaing the date and CO2 concentrations in micro-mol CO2 per mole (ppm), reported on the 2008A SIO manometric mole fraction scale.  These weekly values have been adjusted to 12:00 hours at middle day of each weekly period as indicated by the date in the first column.

You can find the first few lines of the weekly measurement dataset on the following table.

```{r}
head(weekly_co2_consantration)
```

## Observation of the Global Trend and Seasonal trend

When we plot the weekly measurement data set we can easily observe the [Keeling Curve](https://en.wikipedia.org/wiki/Keeling_Curve).
CO2 Concantration in the Atmosphere is increasing by years.

```{r}
library(ggplot2) 

ggplot(weekly_co2_consantration, aes(date, consantration)) + 
    geom_line() + theme_bw() + 
    ggtitle("CO2 Consantration with Time") + 
    theme(plot.title = element_text(hjust = 0.5))

```

When we look at the above graph, there is two trend. One is seasonal trend which makes a cycle and the other is global trend which CO2 concantration always increases with time.

When we look closer to the data there is a cycle related CO2 Concantration according to seasons. Following graph draws the CO2 concentration of 2015, 2016 ans 2017 to observe the seasonal trend better.

```{r}
#Filters data points of the year 2017
weekly_co2_consantration_2017 = subset(weekly_co2_consantration, 
                                       grepl("2015", weekly_co2_consantration$date, fixed=TRUE) | 
                                       grepl("2016", weekly_co2_consantration$date, fixed=TRUE) |
                                       grepl("2017", weekly_co2_consantration$date, fixed=TRUE) )

ggplot(weekly_co2_consantration_2017, aes(date, consantration)) + 
  geom_line() + theme_bw() + 
  ggtitle("CO2 Consantration in 2015-16-17") + 
  theme(plot.title = element_text(hjust = 0.5))

```

## Selection of a model to predict future CO2 concantration

I used anova function of R to compare the performance of models interms of how the model explains varience. I finished the model creation when the new model does not explain extra variance.

I have created 7 models. *m1* is the linear model and others(m2 to m7) are the higher order polinomial models. With the help of anova function I saw that adding 7 term did not contribute to explain variance in the data so I have chosed model 6 as my final model to predict the future.


```{r}
library(ggplot2) 
library("dplyr")

#We can not use date field in calculations so I am creating another field to represent date field
weekly_co2_consantration <- weekly_co2_consantration %>% mutate(index = row_number())
weekly_co2_consantration <- weekly_co2_consantration %>% mutate(index2 = row_number()^2)
weekly_co2_consantration <- weekly_co2_consantration %>% mutate(index3 = row_number()^3)
weekly_co2_consantration <- weekly_co2_consantration %>% mutate(index4 = row_number()^4)
weekly_co2_consantration <- weekly_co2_consantration %>% mutate(index5 = row_number()^5)
weekly_co2_consantration <- weekly_co2_consantration %>% mutate(index6 = row_number()^6)
weekly_co2_consantration <- weekly_co2_consantration %>% mutate(index7 = row_number()^7)

#Creation of models
m1 <- lm(consantration ~ index, data=weekly_co2_consantration )
m2 <- lm(consantration ~ (index + index2), data=weekly_co2_consantration )
m3 <- lm(consantration ~ (index + index2 + index3), data=weekly_co2_consantration )
m4 <- lm(consantration ~ (index + index2 + index3 + index4), data=weekly_co2_consantration )
m5 <- lm(consantration ~ (index + index2 + index3 + index4 + index5 ), data=weekly_co2_consantration )
m6 <- lm(consantration ~ (index + index2 + index3 + index4 + index5 + index6 ), data=weekly_co2_consantration )
m7 <- lm(consantration ~ (index + index2 + index3 + index4 + index5 + index6 + index7), data=weekly_co2_consantration )

#Comparison of models
anova(m1,m2,m3,m4,m5,m6,m7)

#Making Prediktion
m6_predict <- data.frame(consantration = predict(m6, weekly_co2_consantration), date=weekly_co2_consantration$date)


#Observation of the performance of the selected model
ggplot(weekly_co2_consantration, aes(date, consantration)) + 
    geom_line(color="gray") + 
    geom_line(color="blue",data =m6_predict, aes(date, consantration)) + 
    theme_bw() +
    ggtitle("CO2 Consantration with Time and Selected model") + 
    theme(plot.title = element_text(hjust = 0.5))



```


##Prediction of CO2 concantration to 2025

```{r}
library(lubridate)

#Creation of new dates to make future prediction
lastDate =  tail(weekly_co2_consantration, 1)$date
while (!grepl("2050", lastDate, fixed=TRUE)) {
  lastDate <- lastDate + 7
  weekly_co2_consantration[nrow(weekly_co2_consantration) + 1,] = list(lastDate,0)
}

```


```{r}
library(ggplot2) 
library("dplyr")

weekly_co2_consantration <- weekly_co2_consantration %>% mutate(index = row_number())
weekly_co2_consantration <- weekly_co2_consantration %>% mutate(index2 = row_number()^2)
weekly_co2_consantration <- weekly_co2_consantration %>% mutate(index3 = row_number()^3)
weekly_co2_consantration <- weekly_co2_consantration %>% mutate(index4 = row_number()^4)
weekly_co2_consantration <- weekly_co2_consantration %>% mutate(index5 = row_number()^5)
weekly_co2_consantration <- weekly_co2_consantration %>% mutate(index6 = row_number()^6)
weekly_co2_consantration <- weekly_co2_consantration %>% mutate(index7 = row_number()^7)

#Making Prediktion
prediction_linear <- data.frame(consantration = predict(m1, weekly_co2_consantration), date=weekly_co2_consantration$date)
prediction_quadratic <- data.frame(consantration = predict(m2, weekly_co2_consantration), date=weekly_co2_consantration$date)
prediction_initialy_selected <- data.frame(consantration = predict(m6, weekly_co2_consantration), date=weekly_co2_consantration$date)

prediction3 <- data.frame(consantration = predict(m3, weekly_co2_consantration), date=weekly_co2_consantration$date)
prediction4 <- data.frame(consantration = predict(m4, weekly_co2_consantration), date=weekly_co2_consantration$date)
prediction5 <- data.frame(consantration = predict(m5, weekly_co2_consantration), date=weekly_co2_consantration$date)

color_brown = "#CC7600"
color_green = "#029900"
color_yellow = "#FFEB00"
color_rose = "#FF8C9C"
color_red = "#CC1F00"

#Observation of the performance of the selected model
ggplot(weekly_co2_consantration, aes(date, consantration)) + 
    geom_line(color="gray") +
    geom_line(color=color_red,data =prediction_linear, aes(date, consantration)) + 
    geom_line(color=color_yellow,data =prediction_quadratic, aes(date, consantration)) + 
    geom_line(color=color_rose,data =prediction_initialy_selected, aes(date, consantration)) + 
    geom_line(color="blue",data =prediction3, aes(date, consantration)) + 
    geom_line(color=color_green,data =prediction4, aes(date, consantration)) + 
    geom_line(color=color_brown,data =prediction5, aes(date, consantration)) + 
    theme_bw() +
    ggtitle("CO2 Consantration with Time and Selected model") + 
    theme(plot.title = element_text(hjust = 0.5))

```

